\paperTitle

### 2014-01-31 doc/icml2014/directed.tex
more changes, especially to asymptotic variance

@@

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$, allowing us to use
  \TensorFactorize, as reviewed in \sectionref{setup}, to recover
  $O$.

@@

\subsection{General algorithm}
\label{sec:directedGeneral}

Note that in each of the above cases,
each hidden variable $h_i$ of a clique has its own view which is conditionally independent
given the other variables in the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
  A clique $\sC$ of $\sG$ is said to possess \textbf{exclusive views property} if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$.
  and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

\paragraph{Learning cliques}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify unique views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}

\paragraph{Structural properties}

@@

\todo{somewhere give examples that are not in our class, preferably the simplest one, cite Sontag}

@@

      \FOR{each hidden variable $h_i \in H$} 
        \STATE Apply \TensorFactorize to learn conditional moments
        $\mOpp{v}{i}$ for every $\sV_{h_i}$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$} 
  \STATE Apply \LearnClique to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

\paragraph{Consistency and Sample Complexity}

\LearnMarginals combines two consistent algorithms, \TensorFactorize and
\LearnCliqueNs, and is thus consistent itself.
%\todo{did they prove this; if so just say "they show", not "we can show"}
%Using results from
%  \citet{anandkumar12moments,anandkumar13tensor} we can show that
%  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
%  x_{v_2}, x_{v_3}$ has the following sample complexity,
%  \todo{define hat O}
%  \todo{max subscript is misaligned}
%\begin{align*}
%  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O\left( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} \right). 
%\end{align*}

\todo{call MOM squared}
We next describe the asymptotic variance for \LearnClique when using the
  method of moments.
\begin{lemma}(Asymptotic variance of $Z_\sC$)
  \label{lem:mom-variance}  
  The asymptotic variance of $Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit,
  \end{align*}
  where $\pinv{\mOppAll} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
  \cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $d^m \times k^m$ matrix, and 
\begin{align*}
  \Sigma_\sV &= \text{\rm diag}(M_\sV) - M_\sV M_\sV^\top,
\end{align*}
  is the asymptotic variance of the observations $M_\sV$.
\end{lemma}
\begin{proof}
We have for each clique,
\begin{align*}
  Z_{\sC} &= \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m}) \\
\end{align*}

%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation as:
\begin{align*}
  Z_{\sC} &= \pinv{\mOppAll} \mO_\sV.
\end{align*}

From classical asymptotics \cite{vaart98asymptotic},
we have that the asymptotic variance of $M_\sV$ is:
\begin{align*}
  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
\end{align*}
where $\Sigma_\sV$ is the variance of the observations. 
Finally, we have that $Z_\sC$ goes as
\begin{align*}
  \sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \pinvt{\mOppAll} \Sigma_\sV \pinv{\mOppAll}).
\end{align*}
\end{proof}

>>>>

\todo{we are using clique to refer to only hidden cliques; be explicit about using "hidden"}

@@

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$, allowing us to use
  $\TensorFactorize$, as reviewed in \sectionref{setup}, to recover
  $O$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{General algorithm}
\label{sec:directedGeneral}

Note that in each of the above cases,
each hidden variable $h_i$ of the clique has its own view which is conditionally independent
given the other variables in the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
  A clique $\sC$ of $\sG$ is said to possess \textbf{exclusive views property} if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$.
  and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Estimating hidden clique marginals}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \todo{section}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Structural properties}

@@

\todo{somewhere give examples that are not in our class, preferably the simplest one - noisy-or, cite Sontag}

@@

      \FOR{each hidden variable $h_i \in H$} 
        \STATE Apply $\TensorFactorize$ to learn conditional moments
        $\mOpp{v}{i}$ for every $\sV_{h_i}$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$} 
\STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample complexity}

% PL: "combine" is too vague to be useful
%\LearnMarginals combines two consistent algorithms, \TensorFactorize and
%\LearnCliqueNs, and is thus consistent itself.

In this section, we provide formal conditions under which $\LearnMarginals$
will produce consistent estimates.
We let
$\mOpphat{v}{i}$,
$\hat Z_\sC$,
and $\hat M_\sV$,
denote estimators for
$\mOpp{v}{i}$,
$Z_\sC$,
and $M_\sV$,
respectively.

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full column rank and that $\pi^{(i)} \succ 0$ \todo{define $\pi$}.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
every order-1 slice of the marginals $Z_\sC$ has full column rank $k$.
\end{assumption}
According to our directed grid example \todo{ref}, with $\sC = \{ h_1, h_2 \}$, then this condition implies
$\pi = \bP(h_1) \succ 0$ and 
$T = \bP(h_2 \mid h_1) \succ 0$;
with $\sC = \{ h_1, x_1 \}$, we get
$O = \bP(x_1 \mid h_1) \succ 0$.

%\todo{did they prove this; if so just say "they show", not "we can show"}
%Using results from
%  \citet{anandkumar12moments,anandkumar13tensor} we can show that
%  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
%  x_{v_2}, x_{v_3}$ has the following sample complexity,
%  \todo{define hat O}
%  \todo{max subscript is misaligned}
%\begin{align*}
%  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O\left( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} \right). 
%\end{align*}

From \cite{anandkumar13tensor} given \assumptionref{full-rank-plus},
we have that $\mOpphat{v}{i}$ converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant that
depends polynomially on the $k$-th singular value of $\mOpp{v}{i}$.
Of course, the singular value of $\mOpp{v}{i}$ become exponentially worse as $x_v$ and $h_i$ become farther.
So while we can obtain consistent estimates, the sample complexity
is exponential in the mimimum distance from $h_i$ to any three views, as to be expected.
%(\todo{say something
%quantitative?  annoying part is that we would have to bring in the bounds, so leave it out probably}).

% DONE: need to setup the asymptotic - it's not fresh in everyone's mind.
% Easy case
Next, we need to show that the hidden marginals $\hat Z_\sC$ converge to $Z_\sC$.
First, $\hat M_\sV$ is just an empirical average of multinomials over the data points.
Abusing notation slightly, we let $M_\sV$ also denote its $d^{|\sV|}$-dimensional vectorized version.
We also represent $\mH_\sC$ as
  a vector in $\Re^{k^m}$, and represent $\mOppAll \eqdef
  \mOpp{v_1}{i_1} [x_{v_1}] \otimes \cdots \otimes
  \mOpp{v_m}{i_m} [x_{v_m}]$ as a matrix in $\Re^{d^m \times
  k^m}$.
By the central limit theorem, we have:
$\sqrt{n} (\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV)$,
where $\Sigma_\sV$ is the \emph{asymptotic variance} of $\hat M_\sV$ \todo{doesn't this need to be multiplied by something?}:
\begin{align*}
\Sigma_\sV \eqdef \dM_\sV - M_\sV M_\sV^\top, \quad \dM_\sV \eqdef \text{diag}(M_\sV).
\end{align*}

We can now use the delta-method to convert the above result into 
the asymptotic variance for the pseudoinverse version of $\LearnClique$:
\begin{lemma}[Asymptotic variance of pseudoinverse estimator for $Z_\sC$]
  \label{lem:mom-variance}  
  Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
  Then the asymptotic variance of $Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit,
  \end{align*}
  where $\pinv{\mOppAll} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
  \cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $d^m \times k^m$ matrix.
\end{lemma}
\begin{proof}
For clique $\sC$, recall we have
  $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation more compactly as: $Z_{\sC} = \pinv{\mOppAll} \mO_\sV$.
By the delta-method \cite{vaart98asymptotic},
% DONE: need to shorten
%we have that the asymptotic variance of $M_\sV$ is:
%\begin{align*}
%  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
%\end{align*}
%where $\Sigma_\sV$ is the variance of the observations. 
we immediately get:
$\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
\end{proof}
